---
title: "Data QA/QC"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

## Basic setup for plotting and data handling 

```{r, message=F}
library(tidyverse) # Tools for data science (graphing, data reorganizing, etc.)
# Some custom graphing stuff
source("theme_pub.R")
theme_set(theme_pub())
```

```{r}
InDat<-read.csv("./data/RawData.csv")
```

# Fixing data

## Set batch number to factor

```{r}
InDat$Batch.Number<-as.factor(InDat$Batch.Number)
```

## Separate descrptive columns from measurements

```{r}
DescNames<-c("Sample.Name","Batch.Number","Class.name","Sex","Age","CT")
DesCols<-InDat[,names(InDat) %in% DescNames]
DatCols<-InDat[,!names(InDat) %in% DescNames]
```

## Replace non-numeric values

Replace with NA or 0

```{r}
NAs<-c("FAILED") # Strings to replace with missing values
Zeros<-c("< 0|No Peak|#DIV/0") # Strings to replace with 0 (separated by |)

DatCols<-mutate_if(DatCols,
                  is.character,
                  str_replace_all, pattern = NAs, replacement = "NA") 
DatCols<-mutate_if(DatCols,
                  is.character,
                  str_replace_all, pattern = Zeros, replacement = "0")
```

Convert edited data columns from character data back to numeric data

```{r, warning=F}
DatCols<-mutate_if(DatCols,is.character,as.numeric)

#str(Metabolites) #Check data structure
```

## Remove uninformative rows/cols

```{r}
# Remove columns with no usable data (all NA)
keepCols<-names(DatCols[,colSums(is.na(DatCols)) < nrow(DatCols)])
# Remove rows with no usable data (all NA)
keepRows<-rowSums(is.na(DatCols)) < ncol(DatCols)
#
fixDat<-cbind(DesCols,DatCols[,keepCols])[keepRows,]
```

## Missing data (non-zero)

```{r}
# Calculate # missing values for each row and column
RowNA<-rowSums(is.na(fixDat)) 
RowNA[RowNA>0]

ColNA<-colSums(is.na(fixDat))
ColNA[ColNA>0]
```

## Zeros

```{r}
# Calculate # missing values for each row and column
RowZ<-rowSums(fixDat==0,na.rm=T) 
RowZ[RowZ>0]

ColZ<-colSums(fixDat==0,na.rm=T)
ColZ[ColZ>0]
```

# Data Distribution

Inspect distribution of data. File output to `Concentration_Hist.pdf` for inspection, in a folder called `Output` (you may need to create this folder to avoid errors). Note log-transformation of values (x-axis).

```{r}
Concs<-names(select_if(fixDat,is.numeric)) # Columns with concentration (metabolite) data
HistDat<-fixDat %>% gather(Metab,Conc,all_of(Concs))
```


```{r, eval=F}
pdf("./Output/Concentration_Hist.pdf",width=24,height=24)
  qplot(x=log(Conc+1),fill=Batch.Number,data=HistDat,main="Batch") +
    facet_wrap(vars(Metab),scales="free")
  qplot(x=log(Conc+1),fill=Class.name,data=HistDat,main="Class") +
    facet_wrap(vars(Metab),scales="free")
dev.off()
```

Conclusion: Many concentrations are not normally distributed, even after log-transformation. Non-parametric methods are recommended. 

## Example:

Raw data:

```{r}
qplot(fixDat$C18.1,fill=fixDat$Class.name,posit="dodge")
```

Log-transformed:

```{r}
qplot(log(fixDat$C18.1+1),fill=fixDat$Class.name,posit="dodge")
```

See the [bestNormalize](https://cran.r-project.org/web/packages/bestNormalize/vignettes/bestNormalize.html) package in R for transformation options.

# Filter problematic data

## Missing Data

```{r}
ExSample<-25 # Exclude samples with more than this many zeros or missing values
ExMet<-50 # Exclude metabolites with more than this many zeros or missing values

bestDat<-fixDat[RowNA < ExSample & RowZ < ExSample,
                ColNA < ExMet & ColZ < ExMet | colnames(fixDat) == "CT"]

```

## Excluded columns:

```{r}
names(fixDat[,!names(fixDat) %in% names(bestDat)])
```

## Plot missing data

Pre-correction

```{r}
qplot(rowSums(fixDat==0,na.rm=T))+xlab("Zeros/NAs")+ylab("N Samples")
```

Post-correction

```{r}
qplot(rowSums(bestDat==0,na.rm=T))+xlab("Zeros/NAs")+ylab("N Samples")
```

# Output 

Write the filtered data to csv file for analysis.

```{r}
write.csv(bestDat,"./data/FixedData.csv",row.names=F)
```
